---
title: "Fretboard diagram plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fretboard diagram plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(crayon.enabled = TRUE)
sgr_wrap <- function(x, options){
  paste0("<pre class=\"r-output\"><code>", fansi::sgr_to_html(x = htmltools::htmlEscape(x)), "</code></pre>")
}
knitr::knit_hooks$set(output = sgr_wrap)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, error = FALSE, tidy = FALSE, out.width = "100%"
)
library(ggplot2)
library(purrr)
library(gridExtra)
library(tabr)
```


## Overview

While music visualization in `tabr` focuses on leveraging LilyPond to make tablature, some diagrams can also be drawn directly in R using ggplot without any need to involve LilyPond. The `plot_fretboard()` function makes standalone fretboard diagrams in R, independent of the LilyPond sheet music pipeline.

`plot_fretboard()` is a highly specialized function that takes vector inputs for string numbers and fret numbers and maps the combination element-wise to produce a fretboard diagram. Note that `plot_fretboard()` is a developmental function and the interface and arguments it provides may change.

The fretted notes are marked in the diagram using customary large circles.

```{r fb1}
plot_fretboard(string = 6:1, fret = c(0, 2, 2, 0, 0, 0))
```

## Labels and tuning

The fretted notes can be labeled. `labels` can be an arbitrary vector corresponding to the string and fret numbers. For example, you can label each circle with the fingerings used to play chords or scales.

```{r fb2}
plot_fretboard(6:1, c(0, 2, 2, 0, 0, 0), c("G", "U", "I", "T", "A", "R"))
```

If you set `labels = "notes"`, this is a special setting that will label all points with their note names. This can be done automatically because providing string and fret numbers in conjunction with the `tuning` argument gives full information about the notes along the guitar neck. `plot_fretboard()` transposes these internally. This means it will work automatically no matter what arbitrary `tuning` you set. Here is an example that also displays the tuning.

```{r fb3}
string <- c(6, 6, 6, 5, 5, 5, 4, 4, 4, 4, 4, 3, 3, 3, 2, 2, 2, 1, 1, 1)
fret <- c(2, 4, 5, 2, 4, 5, 2, 4, 6, 7, 9, 6, 7, 9, 7, 9, 10, 7, 9, 10)
plot_fretboard(string, fret, "notes", show_tuning = TRUE)
```

When `labels = "notes"`, accidentals are displayed as flats by default. Set `accidentals = "sharp"` to change. 

```{r fb4}
plot_fretboard(string, fret, "notes", show_tuning = TRUE, accidentals =  "sharp")
```

## Limits

X and Y limits are expressed in terms of the number of instrument strings and the span of frets. You can override the fret range that is derived from `fret`:

```{r fb5}
plot_fretboard(string, fret, "notes", fret_range = c(0, 10), show_tuning = TRUE)
```

The number of strings an instrument has is derived from `tuning` and this generalizes the fretboard diagram possibilities further. The tuning now specifies a seven-string guitar. One note has been added on string seven:

```{r fb6}
tuning <- "b1 e2 a2 d3 g3 b3 e4"
plot_fretboard(c(7, string), c(1, fret), "notes", fret_range = c(0, 10), 
               tuning = tuning, show_tuning = TRUE)
```

## Color and faceting

The points (circles) can have border and fill color. The labels can also be colored separately. These arguments can be vectorized. Notice the open Am chord does not really have an open sixth string; it is muted. The zero is still given in order to specify where to notate, but the `mute` argument is given with a logical vector that indicates this entry is muted.

```{r fb7}
am_frets <- c(c(0, 0, 2, 2, 1, 0), c(5, 7, 7, 5, 5, 5))
am_strings <- c(6:1, 6:1)
mute <- c(TRUE, rep(FALSE, 11))

# colors
idx <- c(2, 2, 1, 1, 1, 2, rep(1, 6))
lab_col <- c("white", "black")[idx]
pt_fill <- c("firebrick1", "white")[idx]

plot_fretboard(am_strings, am_frets, "notes", mute, 
               label_color = lab_col, point_fill = pt_fill)
```

`group` can also be used for faceting. However, faceting is still a problematic feature. It may work well enough in cases where the different diagrams span similar frets. The presence of muted notes can also cause issues when faceting. `plot_fretboard()` works best for single-panel plots. Since the function returns a ggplot object, you can always make them separate plots and arrange in a grid layout rather than rely on within-plot faceting.

Note that `plot_fretboard()` accepts character inputs like those used throughout `tabr`:

```{r fb7b}
f <- "0 2 2 1 0 0 0 2 2 0 0 0"
s <- c(6:1, 6:1)
grp <- rep(c("Open E", "Open Em"), each = 6)

# colors
idx <- c(2, 1, 1, 1, 2, 2, 2, 1, 1, 2, 2, 2)
lab_col <- c("white", "black")[idx]
pt_fill <- c("firebrick1", "white")[idx]

plot_fretboard(s, f, "notes", group = grp, fret_range = c(0, 4),
               label_color = lab_col, point_fill = pt_fill)
```

## Orientation

The direction and handedness can also be changed. Diagrams can be vertical or horizontal as well as left- or right-handed.

Here, titles are added to the ggplot objects with `ggtitle`. Of course you can add onto ggplot objects returned by `plot_fretboard()`, but you are limited in what you can add on and must be careful to avoid overriding properties of layers `plot_fretboard()` has already specified.

```{r fb8}
library(ggplot2)
plot_fretboard(string, fret, "notes", label_color = "white", point_fill = "dodgerblue",
               fret_range = c(0, 10), show_tuning = TRUE, horizontal = TRUE) +
  ggtitle("Horizontal")

plot_fretboard(string, fret, "notes", label_color = "white", point_fill = "dodgerblue",
               fret_range = c(0, 10), show_tuning = TRUE, horizontal = TRUE, left_handed = TRUE) +
  ggtitle("Horizontal and left-handed")
```

## Chord diagrams

The previous examples show a mix of using `plot_fretboard()` to make general fretboard diagrams that show scales, arpeggios and other patterns, as well as to produce chord diagrams of specific chords. It is easier to use `plot_chord()` for the latter. It is a wrapper around `plot_fretboard()` that takes a string representing a single chord in the simple fret format shown here.

```{r plot_chord1}
idx <- c(1, 1, 2, 2, 2, 1)
fill <- c("white", "black")[idx]
lab_col <- c("black", "white")[idx]
plot_chord("xo221o", "notes", label_color = lab_col, point_fill = fill)
```

Leading `x` is inferred if there are fewer fret values in the string than there are instrument strings, as inferred from the `tuning` argument. `plot_chord()` takes all the same arguments as `plot_fretboard()` except that it takes `chord` instead of `string` and `fret` and it does not use `mute` because muted notes are indicated with an `x` inside `chord`.

```{r plot_chord2}
plot_chord("0231")
```

Frets are assumed to be single-digit when provided as above. When any two-digit fret value occurs, you must provide the fret values as space- or semicolon-delimited. The latter is like LilyPond format, but a trailing semicolon is ignored if present.

```{r plot_chord3}
plot_chord("10 12 13 11", fret_range = c(10, 14)) # same as "10;12;13;11"
```

In an effort to leave some flexibility to the user, and while these plotting functions remain under development, plotting multiple chords as a chord chart is intentionally handled separately on a case by case basis. Here is an example using a little extra `ggplot2`, `purrr` and `gridExtra`. It also shows how you can tweak some layout settings to size and arrange elements nicely for your `Rmd` file.

```{r plot_chord4, fig.height=6.5}
library(ggplot2)
library(purrr)
library(gridExtra)

chords <- c("02210", "32010", "0232", "133211", "022100")
id <- c("Am", "C", "D", "F", "E")

g <- map2(chords, id, ~{
  plot_chord(.x, "notes", point_size = 8, fret_range = c(0, 4), accidentals = "sharp", asp = 1.25) + 
    ggtitle(.y)
})

grid.arrange(grobs = g, nrow = 2)
```

You can use `marrangeGrob()` instead of `grid.arrange()` if you want to split a larger set of chord diagrams onto multiple pages.

```{r cleanup, include=FALSE}
detach("package:purrr", unload = TRUE)
```
