---
title: "Non-standard tuning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-standard tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, error = FALSE, tidy = TRUE, out.width = "100%"
)
library(tabr)
tabr_options(lilypond = "C:/Program Files (x86)/LilyPond/usr/bin/lilypond.exe")
```

This section covers non-standard tunings. Different tunings can be associated with individual music staves and are therefore specified when calling `track`. The `tuning` argument to track accepts a variety of predefined commonly used string tunings, defaulting to `standard`.

More importantly, track tuning can also be expressed using explicit pitch for completely arbitrary tunings with any number of strings (`tabr` technically supports up to seven strings as a general rule). While the predefined set is convenient, it is not necessary to use and this is where the use of alternate tunings really shines.

*This and subsequent tutorial sections will be continued/added soon...*

```{r cleanup, echo=FALSE}
unlink("*.mid")
```
