---
title: "Musical scales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Musical scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, error = FALSE, tidy = FALSE, out.width = "100%"
)
library(tabr)
```

Programming with musical scales can be assisted by a collection of functions in `tabr` that check and manipulate musical scales, modes and key signatures.

## Key signatures

For key signatures, `keys` lists valid key signature abbreviation strings for each key as used in `tabr`. There are several `key_is*` functions that return a logical result regarding properties of keys:

* key_is_natural
* key_is_sharp
* key_is_flat
* key_is_major
* key_is_minor

There are also the functions, `key_n_flats` and `key_n_sharps`, that give the number of respective accidentals in a key signature.

```{r key}
keys()

key_is_flat("f")
key_n_flats("f")
```

The previous section gave an overview of noteworthy strings. While some of the functions that help enforce proper notation in R seemed like they did not offer much utility in terms of direct use, it was clear that they were integral to the robustness of other `tabr` functions. Now, looking at these key signature helpers, it may be tempting to dismiss their utility even more quickly because a trained musician does not need to invoke them at the command line to know what result they will return.

These functions are not provided to answer basic questions in an interactive R session so much as they are for programming. These are some of the initial building blocks on top of which more complex functions are built, including many functions in `tabr`. As the collection of music programming helper functions in `tabr` grows, it becomes easy to do more with less.

## Scales

Several predefined musical scales are provided and accessible by calling various `scale_*` functions. 

* scale_chromatic
* scale_diatonic
* scale_minor
* scale_major
* scale_harmonic_minor
* scale_hungarian_minor
* scale_melodic_minor
* scale_jazz_minor



```{r scales}
scale_hungarian_minor(key = "am", collapse = TRUE)
```

You can specify whether the vector result should be returned as is, for convenient vectorized programming pipelines, or collapsed to a single string in keeping with the space-delimited time syntax format common throughout `tabr`. Many functions in `tabr` accept both formats as inputs and/or offer them as outputs. 

You can also specify if octave numbering should be included or stripped. Octave numbering is included by default because this maintains pitch order when the scale does not start on `C`. Every note in a noteworthy string has an implicit octave-3 position if not explicitly stated. Octave numbering attempts to be somewhat balanced around `C3`. If the result is not what is desired, it can be shifted by 12 semitones with `transpose`.

```{r scales2}
scale_major("f", TRUE, ignore_octave = TRUE)
scale_major("f", TRUE, ignore_octave = FALSE)
```

See `help("scale-helpers")` for details. Depending on the scale, other arguments are available.

## Musical modes

The seven modern modes are also available with `mode_*` functions or through `modern_mode` by passing the mode string name. Other functions include `is_mode` and `rotate_mode`.

* mode_ionian
* mode_dorian
* mode_phrygian
* mode_lydian
* mode_mixolydian
* mode_aeolian
* mode_locrian

```{r modes}
modes()
mode_aeolian("c")
```

## Diatonic scale chords

The `scale_chords` function provides a list of diatonic scale chords based on the root note and scale chosen. It returns triads by default and can also return seventh chords.

```{r scale_chords}
scale_chords("b_", "major", "seventh", collapse = TRUE)
scale_chords("f#", "minor", "triad", collapse = TRUE)
```

## Scale degrees

The functions `scale_note` and `scale_degree` map between notes and degree in a given scale. For chords in a noteworthy string, only the root note is considered. For `scale_degree`, if a note is not diatonic to the scale, `NA` is returned. For `scale_note`, degrees outside the range of the scale return `NA`. 

`note_in_scale` can be used to check whether a note is diatonic to a scale. Unlike the other two functions, it strictly accepts single notes. This is more strict that a noteworthy check, which permits chords.

```{r scale_degrees}
x <- "c e gb'd'"
scale_degree(x)
scale_degree(x, key = "a")
scale_degree(x, key = "am")
scale_degree(x, scale = "chromatic")

scale_note(1:7, "d")
scale_note(c(1:8), "dm", "harmonic minor")

note_in_scale("a_ b c", "a_", ignore_octave = TRUE)
```

Other functions in `tabr` also work with scales and some build upon functions introduced here.
