---
title: "Fretboard diagram plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fretboard diagram plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(crayon.enabled = TRUE)
sgr_wrap <- function(x, options){
  paste0("<pre class=\"r-output\"><code>", fansi::sgr_to_html(x = htmltools::htmlEscape(x)), "</code></pre>")
}
knitr::knit_hooks$set(output = sgr_wrap)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, error = FALSE, tidy = FALSE, out.width = "100%"
)
library(tabr)
```


## Overview

While music visualization in `tabr` focuses on leveraging LilyPond to make tablature, some diagrams can also be drawn directly in R using ggplot without any need to involve LilyPond. The `fretboard_plot` function makes standalone fretboard diagrams in R, independent of the LilyPond sheet music pipeline.

`fretboard_plot` is a highly specialized function that takes vector inputs for string numbers and fret numbers and maps the combination element-wise to produce a fretboard diagram. Note that `fretboard_plot` is a developmental function and the interface and arguments it provides may change.

The fretted notes are marked in the diagram using customary large circles.

```{r fb1}
fretboard_plot(string = 6:1, fret = c(0, 2, 2, 0, 0, 0))
```

## Labels and tuning

The fretted notes can be labeled. `labels` can be an arbitrary vector corresponding to the string and fret numbers. For example, you can label each circle with the fingerings used to play chords or scales.

```{r fb2}
fretboard_plot(6:1, c(0, 2, 2, 0, 0, 0), c("G", "U", "I", "T", "A", "R"))
```

If you set `labels = "notes"`, this is a special setting that will label all points with their note names. This can be done automatically because providing string and fret numbers in conjunction with the `tuning` argument gives full information about the notes along the guitar neck. `fretboard_plot` transposes these internally. This means it will work automatically no matter what arbitrary `tuning` you set. Here is an example that also displays the tuning.

```{r fb3}
string <- c(6, 6, 6, 5, 5, 5, 4, 4, 4, 4, 4, 3, 3, 3, 2, 2, 2, 1, 1, 1)
fret <- c(2, 4, 5, 2, 4, 5, 2, 4, 6, 7, 9, 6, 7, 9, 7, 9, 10, 7, 9, 10)
fretboard_plot(string, fret, "notes", show_tuning = TRUE)
```

How accidentals are displayed when `labels = "notes"`, as sharps or flats, can be controlled by setting the `key` to a specific key or simply to `sharp` or `flat`. Here, the key of F contains a flat, which alters the labeling from the previous diagram. 

```{r fb4}
fretboard_plot(string, fret, "notes", show_tuning = TRUE, key = "f")
```

## Limits

X and Y limits are expressed in terms of the number of instrument strings and the span of frets. You can override the fret range that is derived from `fret`:

```{r fb5}
fretboard_plot(string, fret, "notes", fret_range = c(0, 10), show_tuning = TRUE)
```

The number of strings an instrument has is derived from `tuning` and this generalizes the fretboard diagram possibilities further. The tuning now specifies a seven-string guitar. One note has been added on string seven:

```{r fb6}
tuning <- "b1 e2 a2 d3 g3 b3 e4"
fretboard_plot(c(7, string), c(1, fret), "notes", fret_range = c(0, 10), 
               tuning = tuning, show_tuning = TRUE)
```

## Color and faceting

The points (circles) can have border and fill color. The labels can also be colored separately. These arguments can be vectorized. Notice the open Am chord does not really have an open sixth string; it is muted. The zero is still given in order to specify where to notate, but the `mute` argument is given with a logical vector that indicates this entry is muted.

```{r fb7}
am_frets <- c(c(0, 0, 2, 2, 1, 0), c(5, 7, 7, 5, 5, 5))
am_strings <- c(6:1, 6:1)
mute <- c(TRUE, rep(FALSE, 12))

# colors
idx <- c(2, 2, 1, 1, 1, 2, rep(1, 6))
lab_col <- c("white", "black")[idx]
pt_fill <- c("firebrick1", "white")[idx]

fretboard_plot(am_strings, am_frets, "notes", mute, 
               label_color = lab_col, point_fill = pt_fill)
```

`group` can also be used for faceting. However, faceting is still a problematic feature. It may work well enough in cases where the different diagrams span similar frets. The presence of muted notes can also cause issues when faceting. `fretboard_plot` works best for single-panel plots. Since the function returns a ggplot object, you can always make them separate plots and arrange in a grid layout rather than rely on within-plot faceting.

Note that `fretboard_plot` accepts character inputs like those used throughout `tabr`:

```{r fb7b}
f <- "0 2 2 1 0 0 0 2 2 0 0 0"
s <- c(6:1, 6:1)
grp <- rep(c("Open E", "Open Em"), each = 6)

# colors
idx <- c(2, 1, 1, 1, 2, 2, 2, 1, 1, 2, 2, 2)
lab_col <- c("white", "black")[idx]
pt_fill <- c("firebrick1", "white")[idx]

fretboard_plot(s, f, "notes", group = grp, fret_range = c(0, 4),
               label_color = lab_col, point_fill = pt_fill)
```

## Orientation

The direction and handedness can also be changed. Diagrams can be vertical or horizontal as well as left- or right-handed.

Here, titles are added to the ggplot objects with `ggtitle`. Of course you can add onto ggplot objects returned by `fretboard_plot`, but you are limited in what you can add on and must be careful to avoid overriding properties of layers `fretboard_plot` has already specified.

```{r fb8}
library(ggplot2)
fretboard_plot(string, fret, "notes", label_color = "white", point_fill = "dodgerblue",
               fret_range = c(0, 10), show_tuning = TRUE, horizontal = TRUE) +
  ggtitle("Horizontal")

fretboard_plot(string, fret, "notes", label_color = "white", point_fill = "dodgerblue",
               fret_range = c(0, 10), show_tuning = TRUE, horizontal = TRUE, left_handed = TRUE) +
  ggtitle("Horizontal and left-handed")
```
